The role of advection versus diffusion in intraventricular transport

### 2025-03-20 paper/main.tex
update output directory structure
results, intro

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental endeavors are challenging to pursue due to
technical limitations: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they failed to identify the role of other contributions to CSF
flow and accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF movement in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive alternative to experimental
studies. Numerous previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004Cilia-relatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution. In summary, our work presents a computational
framework together that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.

@@

We model the net forces induced by motile cilia as a steady
bidirectional traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}). Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($\fixme{26.5}$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, \fixme{g}}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{\fixme{f}}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $\fixme{4.8}$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

When combining the contributions, we recognize flow features
persisting throughout the cardiac cycle, now with vortex structures in
the anterior and middle ventricles and directional flow in the
posterior ventricle (\Crefpanels{fig:fig1}{h--j}, \fixme{Supplementary
  video 1}). These characteristic patterns suggest that the flow in
the anterior and middle ventricles is primarily driven by the motile
cilia, while the cardiac pressure pulsations dominate within the
posterior ventricle and in the ducts connecting the ventricular
compartments. The CSF flow speed peaks at the time of peak caudal
cardiac pulsatile flow, when both the tangential traction induced by
the cilia and the pulsatile pressure-driven flow are aligned with the
rostrocaudal axis, reaching up to 28.0 \textmu m/s dorsally in the
posterior part of the middle ventricle. The flow Reynolds number is
$\mathrm{Re} \approx 0.004$, justifying the assumption of Stokes
flow. In summary, the computational CSF flow model reproduces the flow
features observed in vivo~\cite{Olstad2019CiliaryDevelopment}.

@@

Proteins, neurotransmitters and other molecular substances
intrinsically diffuse within the brain ventricles, but are also
advected by the flow of CSF. A natural question is which of these
mechanisms that dominate for any given type of solute. To address this
question, we simulate the distribution and evolution of three
different solutes injected within a region of interest (ROI1) in the
dorsal region of the middle ventricle (\Crefpanels{fig:fig2}{a}),
subject to the CSF flow induced by motile cilia and cardiac pressure
pulsations. In particular, we consider three different molecular
diffusion coefficients, corresponding to a large extracellular vesicle ($D_1 = 2.17
\times 10^{-12}$ m$^2$/s), a 90 kDa protein STM-GFP ($D_2 = 5.75
\times 10^{-11}$ m$^2$/s), and a 26 kDa photoconvertible protein
Dendra2 ($D_3 = 1.15 \times 10^{-10}$ m$^2$/s).

@@

\begin{figure}%[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure2_simulation_results_different_coeffs.png}\caption{Simulated distribution and evolution after photoconversion of
    extracellular vesicles (EVs) ($D_1$), STM-GFP ($D_2$) and Dendra2 ($D_3$). 
    \textbf{a.} Geometry-centered clip of the ventricles mesh showing the locations of the ROIs.
    \textbf{b--d.} Slice of the concentration field simulated
    with diffusion coefficients $D_1$, $D_2$ and $D_3$ for eight time instants covering the
    time range from 45 seconds to 790 seconds, corresponding to 100 to 1750 cardiac cycles.
    Slices are taken in the $xz$-plane ($y=0.140$ mm, the center of the geometry).
    The colorbar is scaled so the maximum is the maximum concentration of the Dendra2 
    concentration field slice at the last time instant.
    \textbf{e--g.} Mean concentration $\cbar$ in each region of interest (ROI)
    as function of time for diffusion coefficient $D_1$, $D_2$ and $D_3$, respectively. 
    \textbf{h.} The time $\hat{t}$ before the mean concentration $\cbar$ exceeds a
    threshold value of $0.25$ (ROIs 2-4) or $0.10$ (ROIs 5, 6) in each simulation setup.
    Note that there is no bar for $D_1$ in ROI 6, because the threshold value was never exceeded.}\label{fig:fig2}
\end{figure}

To further quantify the solute transport, we examine the mean
concentrations $\cbar_i(t)$ in the six ROIs over time
(\Crefpanels{fig:fig2}{e--g}) and, in addition, the times $\hat{t}_i$
when the mean concentrations first exceed a threshold value
$\hat{c}_i$ (\Crefpanels{fig:fig2}{h}).  Higher diffusion coefficients
result in more rapid transport (\Crefpanels{fig:fig2}{e--g}), with
lower times-to-threshold in ROIs 2--6 (\Crefpanels{fig:fig2}{h}). For
all three diffusion coefficients, the solute first spreads within the
middle ventricle, covering ROIs 2--4, before appearing in the anterior
ventricle (ROI5) and finally reaching the posterior ventricle
(ROI6). Notably, the extracellular vesicles, associated with the smallest diffusion
coefficient, do not reach the posterior ventricle (ROI6) within the
simulated time frame.

@@

Comparing the intensity changes with the simulated concentration
profiles in ROIs 1--6 (\Crefpanels{fig:fig3}{a, b}) show that the
model predictions agree well with the experimental measurements. We
note minor differences however. The simulated values of $\cbar_i$ at
the final time (0.72, 0.82, 0.56 and 0.35 in ROIs 2--6, respectively)
are generally higher than the corresponding experimental $\Delta F$
values (0.63, 0.71, 0.67, 0.29 and 0.18). Moreover, we observe
differences in the distribution within the middle ventricle,
especially with regard to ROIs 2 and 3. Overall, we note that the time
to reach a 25\% threshold agree well between the model predictions and
experimental values (\Crefpanels{fig:fig3}{c, d}).

%% the $\cbar$ profiles of ROIs 3--4 are opposite order in the
%% simulations as compared to the experiments.

%% Notwithstanding similar dynamics, we observe small differences between
%% the experiments and simulation.

%% For instance, transport in ROIs 2--6 is faster in the simulation and
%% reaches higher final values.

@@

\mer{Sentence on why absence of motile cilia is intersting: Cilia
  motion is key to development and absence of cilia is associated with
  pathologies.} As observed both experimentally and computationally,
the absence of motile cilia alters the CSF flow patterns. We next
investigate the implications of these changes on intraventricular
transport.

We first experimentally measured the movement of photoconverted
Dendra2 signal in control and motile-cilia mutant (\emph{smh})
zebrafish. In ROIs 2 and 4, which measure Dendra2 distribution
downstream of the photoconversion site, we observed a slower increase
of fluorescence intensity (\Crefpanels{fig:fig4}{a}), resulting in
$45\%$ (ROI 2) and $25\%$ (ROI 4) higher mean times-to-threshold
(\Crefpanels{fig:fig4}{b}), in the \emph{smh} mutants compared to
controls. We found no significant differences in the Dendra2
distribution in ROIs 3, 5, and 6 between \emph{smh} mutants and
controls.

The \emph{smh} mutants are mimicked in the computational model by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in-silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}).

Compared to experimental data, the difference between the control and
mutant scenarios was even greater for the simulations, with the
times-to-threshold in ROI 2 and 4 being 90\% and 34\% higher in the
absence of cilia-mediated flow. Moreover, the ROI 5 $\cbar$ profiles
for the two configurations in our computational model deviate,
which is not the case in the experiments.  Altogether,
our findings identify that, for smaller proteins such as Dendra2,
diffusion plays an important role for
solute movement to regions upstream in the cilia flow fields or to
distant ventricles, while cilia affect the downstream mean concentrations.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $F(t)$
    in each region of interest (ROI) for control (black lines, $n=8$)
    and \emph{smh} mutant (red lines, $n=6$) fish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard devation errors
    in the means.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for control (gray dots) and
    \emph{smh} mutant (red dots) fish.
    The thick black lines denote mean values of the datasets. The $p$-values
    were calculated using a Mann-Whitney U test with a confidence level of 95\%, with 
    the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-only/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in both panels \textbf{a} and \textbf{c}, the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

We previously characterized three distinct ciliated cell lineages in the
zebrafish larval brain~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain}.
To identify the effects of these lineages on CSF flow and transport, 
we investigate region-specific cilia paralysis with the computational model.
Three cilia-covered regions were considered paralyzed separately:
the telencephalic (anterior) ventricle, and the dorsal and ventral parts
of the diencephalic (middle) ventricle (\Crefpanels{fig:fig5}{a}).
To simulate cilia paralysis, the cilia tangential traction boundary condition
is for each case changed to a free-slip boundary condition
in the paralyzed region,
prescribing zero tangential forces to
represent absence of cilia motion.

The CSF velocities are most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity magnitude drops from \fixme{27.7} \textmu m/s to \fixme{8.0} \textmu m/s.
In vivo, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic cilia has minimal
impact and results in maximum speeds of \fixme{27.7} \textmu m/s or \fixme{27.2} \textmu m/s, respectively.
Of the total tangential forces exerted by the cilia, ventral cilia in the middle ventricle
contribute with 34\%, while the dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). The magnitude of the force contributions are thus
more similar than the difference in their impact on the maximum velocity magnitudes.

Changes in the transport of Dendra2 are most prominent when removing the dorsal cilia
in the diencephalic ventricle. The mean concentration profiles 
of all ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit changes (\Crefpanels{fig:fig5}{c}).
Compared to the baseline model results, we observe less solute distribution
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
Meanwhile, removal of telencephalic cilia does not alter the 
mean concentrations (\Crefpanels{fig:fig5}{c}).
Finally, without ventral cilia in the diencephalic ventricle,
more solute spreads towards the anterior ventricle (ROI 5) and
less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} The total tangential force applied with the tangential traction
    boundary condition in the three separate ciliated regions, as well as the total
    (the sum of the three regions), calculated with \Cref{eq:tangential_force_calculation}. 
    \textbf{c.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

\subsection{Modification of ventricular geometry impacts solute distribution}
Variations in ventricular morphology impact CSF flow patterns.
To assess how these variations influence solute distribution,
we consider four alternative brain ventricles meshes:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Crefpanels{fig:fig6}{a, b}). In these
four modified geometries the volume is reduced by 3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively,
compared to the original geometry.

>>>>

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental endeavors are challenging to pursue due to
technical limitations: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they failed to identify the role of other contributions to CSF
flow and the accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF motion in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive alternative to experimental
studies. Numerous previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}, but also 
other applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella}, and
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004CiliaRelatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution. In summary, our work presents a computational
framework that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.

@@

We model the net forces induced by motile cilia as a steady
bidirectional traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}). Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, \fixme{g}}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{\fixme{f}}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

When combining the contributions, we recognize flow features
persisting throughout the cardiac cycle, now with vortex structures in
the anterior and middle ventricles and directional flow in the
posterior ventricle (\Crefpanels{fig:fig1}{h--j}, \fixme{Supplementary
  video 1}). These characteristic patterns suggest that the flow in
the anterior and middle ventricles is primarily driven by the motile
cilia, while the cardiac pressure pulsations dominate within the
posterior ventricle and in the ducts connecting the ventricular
compartments. The CSF flow speed peaks at the time of peak caudal
cardiac pulsatile flow, when both the tangential traction induced by
the cilia and the pulsatile pressure-driven flow are aligned with the
rostrocaudal axis, reaching up to 27.9 \textmu m/s dorsally in the
posterior part of the middle ventricle. The flow Reynolds number is
$\mathrm{Re} \approx 0.004$, justifying the assumption of Stokes
flow. In summary, the computational CSF flow model reproduces the flow
features observed in vivo~\cite{Olstad2019CiliaryDevelopment}.

@@

Proteins, neurotransmitters and other molecular substances
intrinsically diffuse within the brain ventricles, but are also
advected by the flow of CSF. A natural question is which of these
mechanisms that dominate for any given type of solute. To address this
question, we simulate the distribution and evolution of three
different solutes injected within a region of interest (ROI 1) in the
dorsal region of the middle ventricle (\Crefpanels{fig:fig2}{a}),
subject to the CSF flow induced by motile cilia and cardiac pressure
pulsations. In particular, we consider three different molecular
diffusion coefficients, corresponding to a large extracellular vesicle ($D_1 = 2.17
\times 10^{-12}$ m$^2$/s), a 90 kDa protein STM-GFP ($D_2 = 5.75
\times 10^{-11}$ m$^2$/s), and a 26 kDa photoconvertible protein
Dendra2 ($D_3 = 1.15 \times 10^{-10}$ m$^2$/s).

@@

\begin{figure}%[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure2_simulation_results_different_coeffs.png}\caption{Simulated distribution and evolution after photoconversion of
    extracellular vesicles (EVs) ($D_1$), STM-GFP ($D_2$) and Dendra2 ($D_3$). 
    \textbf{a.} Geometry-centered clip of the ventricles mesh showing the locations of the ROIs.
    \textbf{b--d.} Slices of the concentration field simulated
    with diffusion coefficients $D_1$, $D_2$ and $D_3$ for nine time instants covering the
    time range from 90 seconds to 811 seconds, corresponding to 200 to 1800 cardiac cycles,
    for even intervals of 200 cycles.
    Slices are taken in the $xz$-plane ($y=0.140$ mm, the center of the geometry).
    The colorbar is scaled so the maximum is the maximum concentration of the Dendra2 
    concentration field slice at the last time instant.
    \textbf{e--g.} Mean concentration $\cbar$ in each region of interest (ROI)
    as function of time for diffusion coefficients $D_1$, $D_2$ and $D_3$, respectively. 
    \textbf{h.} The time $\hat{t}$ before the mean concentration $\cbar$ exceeds a
    threshold value of $0.25$ (ROIs 2-4) or $0.10$ (ROIs 5, 6) in each simulation setup.
    Note that there is no bar for $D_1$ in ROI 6, because the threshold value was never exceeded.}
    \label{fig:fig2}
\end{figure}

To further quantify the solute transport, we examine the mean
concentrations $\cbar_i(t)$ in the six ROIs over time
(\Crefpanels{fig:fig2}{e--g}) and, in addition, the times $\hat{t}_i$
when the mean concentrations first exceed a threshold value
$\hat{c}_i$ (\Crefpanels{fig:fig2}{h}).  Higher diffusion coefficients
result in more rapid transport (\Crefpanels{fig:fig2}{e--g}), with
lower times-to-threshold in ROIs 2--6 (\Crefpanels{fig:fig2}{h}). For
all three diffusion coefficients, the solute first spreads within the
middle ventricle, covering ROIs 2--4, before appearing in the anterior
ventricle (ROI 5) and finally reaching the posterior ventricle
(ROI 6). Notably, the extracellular vesicles, associated with the smallest diffusion
coefficient, do not reach the posterior ventricle (ROI 6) within the
simulated time frame~\Crefpanels{fig:fig2}{b, e}.

@@

Comparing the intensity changes with the simulated concentration
profiles in ROIs 1--6 (\Crefpanels{fig:fig3}{a, b}) show that the
model predictions agree well with the experimental measurements. We
note minor differences however. The simulated values of $\cbar_i$ at
the final time \fixme{(0.72, 0.82, 0.56, 0.35 and X in ROIs 2--6, respectively)}
are generally higher than the corresponding experimental $\Delta F$
values (0.63, 0.71, 0.67, 0.29 and 0.18). Moreover, we observe
differences in the distribution within the middle ventricle,
especially with regard to ROIs 2 and 3. Overall, we note that the time
to reach a 25\% threshold agree well between the model predictions and
experimental values (\Crefpanels{fig:fig3}{c, d}).

@@

Ciliary motion is integral to physiological CSF flow,
and absence of cilia has been associated with defective
brain development and various pathologies~\cite{Eichele2020Cilia-drivenVentricle,
Olstad2019CiliaryDevelopment, Faubel2016Cilia-basedVentricles,
Guirao2010CouplingCilia, Afzelius2004CiliaRelatedDiseases,
Sawamoto2006NewBrain, Yoshiba2014RolesSymmetry,
Hirokawa2006NodalAsymmetry}.
As observed both experimentally and computationally,
the absence of motile cilia alters the CSF flow patterns. 
We next investigate the implications of these changes on intraventricular
transport.

We first experimentally measured the movement of photoconverted
Dendra2 signal in control and motile-cilia mutant (\emph{smh})
zebrafish. In ROIs 2 and 4, which measure Dendra2 distribution
downstream of the photoconversion site, we observed a slower increase
of fluorescence intensity (\Crefpanels{fig:fig4}{a}). The mean times-to-threshold
in the \emph{smh} mutants were $45\%$ (ROI 2) and $25\%$ (ROI 4) higher
compared to controls (\Crefpanels{fig:fig4}{b}). We found no significant differences
in the Dendra2 distribution in ROIs 3, 5, and 6 between
\emph{smh} mutants and controls.

The \emph{smh} mutants are mimicked in the computational model by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}).

Compared to experimental data, the difference between the control and
mutant scenarios was even greater for the simulations, with the
times-to-threshold in ROI 2 and 4 being 90\% and 34\% higher in the
absence of cilia-mediated flow. Moreover, the ROI 5 $\cbar$ profiles
for the two configurations in our computational model deviate,
which is not the case in the experiments. Altogether,
our findings identify that, for smaller proteins such as Dendra2,
diffusion plays an important role for
solute movement to regions upstream in the cilia-mediated flow fields or to
distant ventricles, while cilia affect the downstream mean concentrations.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $F(t)$
    in each region of interest (ROI) for control (black lines, $n=8$)
    and \emph{smh} mutant (red lines, $n=6$) fish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard devation errors
    in the means.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for control (gray dots) and
    \emph{smh} mutant (red dots) fish.
    The thick black lines denote mean values of the datasets. The $p$-values
    were calculated using a Mann-Whitney U test with a confidence level of 95\%, with 
    the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-only/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in both panels \textbf{a} and \textbf{c}, the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

We previously characterized three distinct ciliated cell lineages in the
zebrafish larval brain~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain}.
To identify the effects of these lineages on CSF flow and transport, 
we investigate region-specific cilia paralysis with the computational model.
Three cilia-covered regions were considered paralyzed separately:
the anterior ventricle, and the dorsal and ventral parts
of the middle ventricle (\Crefpanels{fig:fig5}{a}).
To simulate cilia paralysis, the cilia tangential traction boundary condition
is for each case changed to a free-slip boundary condition
in the paralyzed region,
prescribing zero tangential forces to
represent absence of cilia motion.

The CSF velocities are most strongly
affected when removing cilia in the dorsal part of the middle ventricle,
as the maximum velocity magnitude drops from 27.9 \textmu m/s to 7.9 \textmu m/s.
In vivo, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing cilia in the anterior ventricle or the ventral region of the middle
ventricle has minimal impact, and results in maximum speeds
of 27.9 \textmu m/s and 27.4 \textmu m/s, respectively.
Of the total tangential forces exerted by the cilia, ventral cilia in the middle ventricle
contribute with 34\%, while the dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). The magnitude of the force contributions are thus
more similar than the difference in their impact on the maximum velocity magnitudes.

Changes in the transport of Dendra2 are most prominent when removing the dorsal cilia
in the middle ventricle. All mean concentration profiles 
(except ROI 1, where we impose the photoconversion curve)
exhibit changes (\Crefpanels{fig:fig5}{c}).
Compared to the baseline model results, we observe less solute distribution
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
On the other hand, removal of anterior cilia does not alter the 
mean concentrations (\Crefpanels{fig:fig5}{c}).
Finally, without ventral cilia in the middle ventricle,
more solute spreads towards the anterior ventricle (ROI 5) and
less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} The total tangential force applied with the tangential traction
    boundary condition in the three separate ciliated regions, as well as the total
    (the sum of the three regions), calculated with \Cref{eq:tangential_force_calculation}. 
    \textbf{c.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

\subsection{Modification of ventricular geometry impacts solute distribution}
Variations in ventricular morphology impact CSF flow patterns~\cite{Olstad2019CiliaryDevelopment}.
To assess how these variations influence solute distribution,
we consider four alternative brain ventricles meshes:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Crefpanels{fig:fig6}{a, b}). In these
four modified geometries the volume is reduced by 3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively,
compared to the original geometry.

Constricting the connection between the anterior and middle ventricle reduces
solute transport to the anterior ventricle (ROI 5),
and the mean concentration in ROIs 2--4 and 6 slightly increases, compared to the original geometry
(\Crefpanels{fig:fig6}{c}). This is notwithstanding a
9.5\% reduction in ventral cilia force magnitude
(\Crefpanels{fig:fig6}{d}), which are the cilia that propel CSF and solute towards ROI 6.
Constricting the mid-hind brain connection
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a
shrunk middle ventricle.
Here the transport to the posterior ventricle (ROI 6) is reduced,
in spite of a 6.6\% increase in the ventral cilia forces.
Additionally, for the shrunk middle configuration, the transport upstream of CSF flow towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.
